---
title: "00 — Data Preparation & Leakage-Free Splits"
author: "Mario"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(lubridate)
```

---

## Purpose

## Model Lineage & Design Scope

This notebook is the data layer for **Mario's R experiment only**.
It is intentionally independent from Anthony's Python model.

| | Mario's R model (this repo) | Anthony's Python V1 | Final ensemble / best submission |
|---|---|---|---|
| Language | R / tidymodels + bonsai | Python / LightGBM native | — |
| Min lag horizon | 7 days (anchored at lag-7) | 92 days (= full forecast window) | — |
| Lag set | lag_7/14/21/28/364 | lag_92 … lag_357/364 (38 weekly lags) | — |
| Extra features | unit_price, R09 interactions, Drinks × temp/weekend | — | — |
| Sample weights | 1 + sqrt(quantity) | None | — |
| Specials override | Yes (val-gated) | No | — |

**Lag design note:** Mario's lags are anchored at 7 days, not 92. This is
valid for training and val (Sep 2025) because those rows are isolated splits
with no look-ahead — `lag_7` on a Sep row pulls from Aug, which is history.
It is also fine for test (Oct–Dec) because `00_data_prep.qmd` computes lags
from `raw_train` history only; test features are joined, not computed
on live test-period actuals.

The tradeoff: shorter lags capture more recent demand signal and react to
recent trends faster. The cost is those features are unavailable for
true multi-step production forecasting beyond a 7-day horizon without
recursive filling. For this competition (batch forecast, all history known
through Aug 31), it is leakage-free.

---

## 1. Config

```{r}
#| label: config

# ── Date boundaries ───────────────────────────────────────────────────────────
TRAIN_START <- as.Date("2021-01-01")
TRAIN_END   <- as.Date("2025-08-31")
VAL_START   <- as.Date("2025-09-01")
VAL_END     <- as.Date("2025-09-30")
TEST_START  <- as.Date("2025-10-01")
TEST_END    <- as.Date("2025-12-31")

# ── Paths ─────────────────────────────────────────────────────────────────────
DATA_IN  <- "data/qsr_demand_dataset.csv"
DATA_OUT <- "data/"
```

---

## 2. Load Raw Data

```{r}
#| label: load-raw

df_raw <- read_csv(DATA_IN, show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(restaurant_id, menu_item_id, date)

glue::glue("Rows: {nrow(df_raw)} | Date range: {min(df_raw$date)} to {max(df_raw$date)}")
glue::glue("Missing quantity: {sum(is.na(df_raw$quantity))} ({round(mean(is.na(df_raw$quantity))*100,2)}%)")
```

---

## 3. Raw Splits by Date

Split first — before any cleaning or feature engineering. Val and test
are frozen from this point forward.

```{r}
#| label: raw-splits

raw_train <- df_raw |> filter(date >= TRAIN_START, date <= TRAIN_END)
raw_val   <- df_raw |> filter(date >= VAL_START,   date <= VAL_END)
raw_test  <- df_raw |> filter(date >= TEST_START,  date <= TEST_END)

cat(
  "raw_train:", nrow(raw_train), "rows |",
    sum(is.na(raw_train$quantity)), "null qty\n",
  "raw_val:  ", nrow(raw_val),   "rows |",
    sum(is.na(raw_val$quantity)),   "null qty\n",
  "raw_test: ", nrow(raw_test),  "rows |",
    sum(is.na(raw_test$quantity)),  "null qty\n"
)
```

---

## 4. Calendar Features

These are always safe — every value is known at prediction time.
Computed once on the full dataset so val and test rows get the same columns.

```{r}
#| label: calendar-features

add_calendar_features <- function(df) {
  df |>
    mutate(
      day_of_year       = yday(date),
      day_of_month      = mday(date),
      week_of_year      = isoweek(date),
      quarter           = quarter(date),
      days_in_month     = days_in_month(date),
      days_to_month_end = as.integer(days_in_month - day_of_month),
      days_from_start   = as.integer(date - as.Date("2021-01-01")),
      # Fourier terms — smooth annual seasonality
      sin_doy_1 = sin(2 * pi * 1 * day_of_year / 365.25),
      cos_doy_1 = cos(2 * pi * 1 * day_of_year / 365.25),
      sin_doy_2 = sin(2 * pi * 2 * day_of_year / 365.25),
      cos_doy_2 = cos(2 * pi * 2 * day_of_year / 365.25),
      # Fourier terms — smooth weekly seasonality
      sin_dow   = sin(2 * pi * day_of_week_num / 7),
      cos_dow   = cos(2 * pi * day_of_week_num / 7)
    ) |>
    select(-days_in_month) |>
    # ── Weather breakdown ──────────────────────────────────────────────────────
    mutate(
      temp_bin    = cut(avg_temp_f, breaks = c(-Inf, 32, 55, 75, Inf),
                        labels = c(0, 1, 2, 3)) |> as.numeric(),
      has_precip  = as.integer(precip_inches > 0),
      is_snow     = as.integer(precip_type == "Snow"),
      is_rain     = as.integer(precip_type == "Rain"),
      # ── Interaction features ──────────────────────────────────────────────
      promo_x_weekend = is_promotion      * is_weekend,
      event_x_weekend = is_special_event  * is_weekend,
      # ── Numeric encodings ─────────────────────────────────────────────────
      category_enc    = as.integer(as.factor(category))      - 1L,
      state_enc       = as.integer(as.factor(state))         - 1L,
      day_of_week_enc = as.integer(as.factor(day_of_week))   - 1L,
      restaurant_enc  = as.integer(as.factor(restaurant_id)) - 1L,
      item_enc        = as.integer(as.factor(menu_item_id))  - 1L,
      # ── R09 identity + interactions (R09 is persistent outlier) ───────────
      is_r09             = as.integer(restaurant_id == "R09"),
      r09_x_special_event = as.integer(restaurant_id == "R09") * is_special_event,
      r09_x_promotion    = as.integer(restaurant_id == "R09") * is_promotion,
      # ── Drinks targeted features (highest volume, most weather-reactive) ──
      is_drinks          = as.integer(category == "Drinks"),
      drinks_x_temp      = as.integer(category == "Drinks") * avg_temp_f,
      drinks_x_weekend   = as.integer(category == "Drinks") * is_weekend
    )
}

df_cal <- df_raw |> add_calendar_features()
```

---

## 5. Leakage-Free Lag & Rolling Features

Lags and rolling windows look **backward only** — a lag_7 on Sep 1 pulls
from Aug 25, which is always in the training period. Rolling windows are
anchored at lag-7 so the most recent 7 days are never included.

All rows (train, val, test) get these features computed from the full
sorted series. There is no leakage because `lag()` and `rollmean()`
only see past values by construction.

```{r}
#| label: lag-features

add_lag_features <- function(df) {
  df |>
    arrange(restaurant_id, menu_item_id, date) |>
    group_by(restaurant_id, menu_item_id) |>
    mutate(
      # ── Point lags ──────────────────────────────────────────────────────────
      # Rolling windows are anchored at lag-7 (most recent 7 days excluded).
      # This is leakage-free for training and val rows because those splits are
      # isolated before feature computation. Test rows receive features joined
      # from training history — lag_7 on a test row does NOT see test-period actuals.
      lag_7   = lag(quantity, 7),
      lag_14  = lag(quantity, 14),
      lag_21  = lag(quantity, 21),
      lag_28  = lag(quantity, 28),
      lag_364 = lag(quantity, 364),

      # ── Same day-of-week lags (most meaningful for weekly demand patterns) ──
      same_dow_lag1 = lag(quantity, 7),   # same DOW, 1 week ago
      same_dow_lag2 = lag(quantity, 14),  # same DOW, 2 weeks ago
      same_dow_lag3 = lag(quantity, 21),  # same DOW, 3 weeks ago
      same_dow_mean = (lag(quantity, 7) + lag(quantity, 14) +
                       lag(quantity, 21) + lag(quantity, 28)) / 4,

      # ── Rolling windows anchored at lag-7 (no leakage into recent days) ────
      # shift by 7 first, then roll forward over that shifted series
      roll_mean_7  = lag(quantity, 7) |>
                       (\(x) zoo::rollmeanr(x, k = 7,  fill = NA, na.rm = TRUE))(),
      roll_mean_14 = lag(quantity, 7) |>
                       (\(x) zoo::rollmeanr(x, k = 14, fill = NA, na.rm = TRUE))(),
      roll_mean_28 = lag(quantity, 7) |>
                       (\(x) zoo::rollmeanr(x, k = 28, fill = NA, na.rm = TRUE))(),
      roll_std_7   = lag(quantity, 7) |>
                       (\(x) zoo::rollapply(x, width = 7,  FUN = sd,
                                            fill = NA, align = "right",
                                            na.rm = TRUE))(),
      roll_max_7   = lag(quantity, 7) |>
                       (\(x) zoo::rollapply(x, width = 7,  FUN = max,
                                            fill = NA, align = "right",
                                            na.rm = TRUE))(),
      roll_std_14  = lag(quantity, 7) |>
                       (\(x) zoo::rollapply(x, width = 14, FUN = sd,
                                            fill = NA, align = "right",
                                            na.rm = TRUE))(),
      roll_max_14  = lag(quantity, 7) |>
                       (\(x) zoo::rollapply(x, width = 14, FUN = max,
                                            fill = NA, align = "right",
                                            na.rm = TRUE))(),
      roll_std_28  = lag(quantity, 7) |>
                       (\(x) zoo::rollapply(x, width = 28, FUN = sd,
                                            fill = NA, align = "right",
                                            na.rm = TRUE))(),
      roll_max_28  = lag(quantity, 7) |>
                       (\(x) zoo::rollapply(x, width = 28, FUN = max,
                                            fill = NA, align = "right",
                                            na.rm = TRUE))()
    ) |>
    ungroup()
}

df_feats <- df_cal |> add_lag_features()

cat("Lag/rolling features added. NAs in lag_7:",
    sum(is.na(df_feats$lag_7)), "\n")
```

---

## 6. Seasonal Baseline Feature (`prior_year_avg`)

`prior_year_avg` is the average quantity for the same restaurant × item ×
month × day-of-week, computed **from training rows only**. This is the
strongest single feature from the EDA (dominates LightGBM gain).

It is computed once from `raw_train` and then joined to all rows —
including val and test. No leakage because it only uses history up to
Aug 31 2025.

```{r}
#| label: prior-year-avg

# Compute from training period only (non-null rows)
prior_year_avg <- raw_train |>
  filter(!is.na(quantity)) |>
  mutate(month = month(date)) |>
  group_by(restaurant_id, menu_item_id, month, day_of_week_num) |>
  summarise(prior_year_avg = mean(quantity, na.rm = TRUE), .groups = "drop")

# Combo-level fallback (for month × DOW combos with no training history)
combo_mean <- raw_train |>
  filter(!is.na(quantity)) |>
  group_by(restaurant_id, menu_item_id) |>
  summarise(combo_mean = mean(quantity, na.rm = TRUE), .groups = "drop")

# Join to all rows
df_feats <- df_feats |>
  mutate(month = month(date)) |>
  left_join(prior_year_avg, by = c("restaurant_id", "menu_item_id",
                                    "month", "day_of_week_num")) |>
  left_join(combo_mean,     by = c("restaurant_id", "menu_item_id")) |>
  # Fill any remaining NAs in prior_year_avg with combo_mean
  mutate(prior_year_avg = coalesce(prior_year_avg, combo_mean))

cat("prior_year_avg NAs remaining:", sum(is.na(df_feats$prior_year_avg)), "\n")
```

---

## 7. Canonical Feature List

Every model notebook imports this same list. Adding or removing a feature
means changing it here once — all models inherit the change.

```{r}
#| label: feature-list

FEATURES <- c(
  # Calendar
  "day_of_week_num", "is_weekend", "month", "quarter",
  "day_of_month", "days_to_month_end", "week_of_year", "day_of_year",
  "days_from_start",
  # Fourier — annual and weekly seasonality
  "sin_doy_1", "cos_doy_1", "sin_doy_2", "cos_doy_2",
  "sin_dow",   "cos_dow",
  # Lags
  "lag_7", "lag_14", "lag_21", "lag_28", "lag_364",
  # Rolling means (anchored at lag-7)
  "roll_mean_7", "roll_mean_14", "roll_mean_28",
  # Rolling std — variability / volatility signal
  "roll_std_7", "roll_std_14", "roll_std_28",
  # Rolling max — spike signal
  "roll_max_7", "roll_max_14", "roll_max_28",
  # Same day-of-week
  "same_dow_lag1", "same_dow_lag2", "same_dow_lag3", "same_dow_mean",
  # Seasonal baseline
  "prior_year_avg", "combo_mean",
  # Exogenous — known at prediction time
  "unit_price", "avg_temp_f", "precip_inches",
  "is_holiday", "is_special_event", "is_promotion",
  # Weather breakdown
  "temp_bin", "has_precip", "is_snow", "is_rain",
  # Interaction features
  "promo_x_weekend", "event_x_weekend",
  # Numeric encodings
  "category_enc", "state_enc", "day_of_week_enc",
  "restaurant_enc", "item_enc",
  # R09 identity + interactions
  "is_r09", "r09_x_special_event", "r09_x_promotion",
  # Drinks targeted features
  "is_drinks", "drinks_x_temp", "drinks_x_weekend"
)

TARGET <- "quantity"

cat("Features:", length(FEATURES), "\n")
cat(paste(FEATURES, collapse = ", "))
```

---

## 8. Null Strategies — Training Only

Three variants. Each applies a different strategy to the 0.4% of training
rows with missing quantity, **then** caps outliers at the 99.5th percentile
per combo (also training-only).

Val and test are never modified.

```{r}
#| label: null-strategies

# ── Helper: cap outliers at 99.5th percentile per combo (training only) ───────
cap_outliers <- function(df) {
  df |>
    group_by(restaurant_id, menu_item_id) |>
    mutate(
      cap_995 = quantile(quantity, 0.995, na.rm = TRUE),
      quantity = pmin(quantity, cap_995)
    ) |>
    ungroup() |>
    select(-cap_995)
}

# Split the featured dataset by period
train_feats <- df_feats |> filter(date >= TRAIN_START, date <= TRAIN_END)
val_feats   <- df_feats |> filter(date >= VAL_START,   date <= VAL_END)
test_feats  <- df_feats |> filter(date >= TEST_START,  date <= TEST_END)

# ── Strategy 1: Drop null quantity rows ───────────────────────────────────────
# Cleanest for lag features — no synthetic values ever enter the feature columns.
# lag_364 NaN rows (first year of history) are kept — LightGBM handles NaN
# natively and the "no full-year history" signal is informative, not noise.
train_dropped <- train_feats |>
  filter(!is.na(quantity)) |>
  cap_outliers()

cat("train_dropped:   ", nrow(train_dropped), "rows |",
    sum(is.na(train_dropped$quantity)), "null qty\n")

# ── Strategy 2: Impute with combo median (restaurant × item) ─────────────────
# Median computed from non-null training rows only — no test data used
combo_median_fill <- raw_train |>
  filter(!is.na(quantity)) |>
  group_by(restaurant_id, menu_item_id) |>
  summarise(combo_median = median(quantity, na.rm = TRUE), .groups = "drop")

train_imputed_combo <- train_feats |>
  left_join(combo_median_fill, by = c("restaurant_id", "menu_item_id")) |>
  mutate(quantity = coalesce(quantity, combo_median)) |>
  select(-combo_median) |>
  cap_outliers()

cat("train_imputed_combo:", nrow(train_imputed_combo), "rows |",
    sum(is.na(train_imputed_combo$quantity)), "null qty\n")

# ── Strategy 3: Impute with same day-of-week median per combo ────────────────
# More targeted — respects weekly demand patterns before filling
dow_median_fill <- raw_train |>
  filter(!is.na(quantity)) |>
  group_by(restaurant_id, menu_item_id, day_of_week_num) |>
  summarise(dow_median = median(quantity, na.rm = TRUE), .groups = "drop")

train_imputed_dow <- train_feats |>
  left_join(dow_median_fill,
            by = c("restaurant_id", "menu_item_id", "day_of_week_num")) |>
  mutate(quantity = coalesce(quantity, dow_median)) |>
  select(-dow_median) |>
  cap_outliers()

cat("train_imputed_dow:  ", nrow(train_imputed_dow), "rows |",
    sum(is.na(train_imputed_dow$quantity)), "null qty\n")
```

---

## 9. Verify Val & Test Are Untouched

```{r}
#| label: verify-splits

# Val and test rows come from df_feats — features computed, quantity untouched
cat("val  rows:", nrow(val_feats),  "| qty NAs:", sum(is.na(val_feats$quantity)),  "\n")
cat("test rows:", nrow(test_feats), "| qty NAs:", sum(is.na(test_feats$quantity)), "\n")

# Val should be scoreable — confirm quantity coverage
val_coverage <- 1 - mean(is.na(val_feats$quantity))
cat("Val quantity coverage:", scales::percent(val_coverage, accuracy = 0.01), "\n")

# Test row count must be exactly 69,000
stopifnot("Test must be exactly 69,000 rows" = nrow(test_feats) == 69000)
cat("✓ test has exactly 69,000 rows\n")
```

---

## 10. Write Splits to `data/`

```{r}
#| label: write-splits

write_csv(train_dropped,      paste0(DATA_OUT, "train_dropped.csv"))
write_csv(train_imputed_combo, paste0(DATA_OUT, "train_imputed_combo_median.csv"))
write_csv(train_imputed_dow,   paste0(DATA_OUT, "train_imputed_dow_median.csv"))
write_csv(val_feats,           paste0(DATA_OUT, "val.csv"))
write_csv(test_feats,          paste0(DATA_OUT, "test.csv"))

# Save canonical feature list and config for model notebooks to import
config <- list(
  FEATURES    = FEATURES,
  TARGET      = TARGET,
  TRAIN_END   = as.character(TRAIN_END),
  VAL_START   = as.character(VAL_START),
  VAL_END     = as.character(VAL_END),
  TEST_START  = as.character(TEST_START),
  TEST_END    = as.character(TEST_END)
)
saveRDS(config, paste0(DATA_OUT, "prep_config.rds"))

cat("\n── Files written to data/ ────────────────────────────────\n")
cat("  train_dropped.csv               →", nrow(train_dropped), "rows\n")
cat("  train_imputed_combo_median.csv  →", nrow(train_imputed_combo), "rows\n")
cat("  train_imputed_dow_median.csv    →", nrow(train_imputed_dow), "rows\n")
cat("  val.csv                         →", nrow(val_feats), "rows\n")
cat("  test.csv                        →", nrow(test_feats), "rows\n")
cat("  prep_config.rds                 → feature list + date config\n")
```

---

## 11. How to Use This in a Model Notebook

Every model notebook in `models/` starts with exactly this:

```{r}
# ── Load standardized splits ──────────────────────────────────────────────────
cfg   <- readRDS("data/prep_config.rds")
val   <- read_csv("data/val.csv",   show_col_types = FALSE)
test  <- read_csv("data/test.csv",  show_col_types = FALSE)

# Pick ONE training variant — this is the only line that changes between runs
train <- read_csv("data/train_dropped.csv", show_col_types = FALSE)
# train <- read_csv("data/train_imputed_combo_median.csv", show_col_types = FALSE)
# train <- read_csv("data/train_imputed_dow_median.csv",   show_col_types = FALSE)

FEATURES <- cfg$FEATURES
TARGET   <- cfg$TARGET

# ── wMAPE helper (identical in every model notebook) ─────────────────────────
wmape <- function(actual, predicted) {
  mask <- !is.na(actual) & actual > 0
  sum(abs(actual[mask] - predicted[mask])) / sum(actual[mask])
}
```

When you change the training variant and re-run, the **only variable
that changed** is the null strategy. Any wMAPE difference is attributable
to that choice alone.

---

## Summary

```{r}
#| label: summary-table

tibble(
  file                 = c("train_dropped.csv",
                            "train_imputed_combo_median.csv",
                            "train_imputed_dow_median.csv",
                            "val.csv",
                            "test.csv"),
  null_strategy        = c("Rows with null qty dropped",
                            "Null qty → combo median (rest × item)",
                            "Null qty → same-DOW median per combo",
                            "Unchanged — score on non-null rows only",
                            "Unchanged — all 69,000 rows"),
  rows                 = c(nrow(train_dropped),
                            nrow(train_imputed_combo),
                            nrow(train_imputed_dow),
                            nrow(val_feats),
                            nrow(test_feats)),
  features_from        = c("own history",
                            "own history",
                            "own history",
                            "raw_train history",
                            "raw_train history")
) |>
  print(n = Inf)
```
