---
title: "01 — Baseline Models"
author: "Mario"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(scales)
library(glue)
```

---

## Purpose

Establishes the **honest performance floor** before any ML model runs.
Every model in `models/02_*` and beyond must beat these numbers to justify
its complexity.

Two baselines are evaluated:

| Baseline | Logic |
|---|---|
| `prior_year_avg` | Mean quantity for same restaurant × item × month × DOW from training history |
| `same_dow_mean` | Mean of the 4 most recent same-weekday observations (lags 7/14/21/28) |

Both are already precomputed as feature columns by `00_data_prep.qmd` —
no model fitting required here.

---

## 1. Load Splits

```{r}
#| label: load-splits

cfg  <- readRDS("data/prep_config.rds")
val  <- read_csv("data/val.csv",  show_col_types = FALSE)
test <- read_csv("data/test.csv", show_col_types = FALSE)

FEATURES <- cfg$FEATURES
TARGET   <- cfg$TARGET

# Training variant — baselines don't fit a model, but we load one so the
# pattern is consistent and comparable with downstream notebooks
train <- read_csv("data/train_dropped.csv", show_col_types = FALSE)

glue("train: {nrow(train)} rows | val: {nrow(val)} rows | test: {nrow(test)} rows")
```

---

## 2. wMAPE Helper

This exact function is used in every model notebook. Do not modify it.

```{r}
#| label: wmape-helper

# wMAPE: sum(|actual - predicted|) / sum(actual)
# Only scores rows where actual > 0 and is not NA
wmape <- function(actual, predicted) {
  mask <- !is.na(actual) & actual > 0
  sum(abs(actual[mask] - predicted[mask])) / sum(actual[mask])
}

# Convenience: score a dataframe that has columns `quantity` and `predicted`
score_wmape <- function(df) {
  wmape(df$quantity, df$predicted)
}
```

---

## 3. Baseline 1 — Prior-Year Average

Uses the seasonal average from the same restaurant × item × month ×
day-of-week in prior training years. This is the strongest naive anchor
and the top feature by gain in every LightGBM experiment run so far.

```{r}
#| label: baseline-prior-year

val_b1 <- val |>
  mutate(predicted = coalesce(prior_year_avg, combo_mean)) |>
  mutate(predicted = pmax(predicted, 0))

test_b1 <- test |>
  mutate(predicted = coalesce(prior_year_avg, combo_mean)) |>
  mutate(predicted = pmax(predicted, 0))

wmape_val_b1  <- score_wmape(val_b1)
wmape_test_b1 <- score_wmape(test_b1)

glue("Baseline 1 — prior_year_avg")
glue("  Val  wMAPE (Sep 2025):     {percent(wmape_val_b1,  accuracy = 0.01)}")
glue("  Test wMAPE (Oct-Dec 2025): {percent(wmape_test_b1, accuracy = 0.01)}")
```

---

## 4. Baseline 2 — Same Day-of-Week Mean

Uses the mean of the 4 most recent same-weekday observations. More
reactive to recent demand shifts than `prior_year_avg`, but noisier.

```{r}
#| label: baseline-dow-mean

val_b2 <- val |>
  mutate(predicted = coalesce(same_dow_mean, prior_year_avg, combo_mean)) |>
  mutate(predicted = pmax(predicted, 0))

test_b2 <- test |>
  mutate(predicted = coalesce(same_dow_mean, prior_year_avg, combo_mean)) |>
  mutate(predicted = pmax(predicted, 0))

wmape_val_b2  <- score_wmape(val_b2)
wmape_test_b2 <- score_wmape(test_b2)

glue("Baseline 2 — same_dow_mean")
glue("  Val  wMAPE (Sep 2025):     {percent(wmape_val_b2,  accuracy = 0.01)}")
glue("  Test wMAPE (Oct-Dec 2025): {percent(wmape_test_b2, accuracy = 0.01)}")
```

---

## 5. Baseline 3 — Blended (50/50)

Simple equal-weight blend of both baselines. Often more stable than
either alone because the two signals have different error profiles.

```{r}
#| label: baseline-blend

val_b3 <- val |>
  mutate(
    pred_b1   = coalesce(prior_year_avg, combo_mean),
    pred_b2   = coalesce(same_dow_mean, prior_year_avg, combo_mean),
    predicted = pmax((pred_b1 + pred_b2) / 2, 0)
  )

test_b3 <- test |>
  mutate(
    pred_b1   = coalesce(prior_year_avg, combo_mean),
    pred_b2   = coalesce(same_dow_mean, prior_year_avg, combo_mean),
    predicted = pmax((pred_b1 + pred_b2) / 2, 0)
  )

wmape_val_b3  <- score_wmape(val_b3)
wmape_test_b3 <- score_wmape(test_b3)

glue("Baseline 3 — 50/50 blend")
glue("  Val  wMAPE (Sep 2025):     {percent(wmape_val_b3,  accuracy = 0.01)}")
glue("  Test wMAPE (Oct-Dec 2025): {percent(wmape_test_b3, accuracy = 0.01)}")
```

---

## 6. Scorecard

```{r}
#| label: scorecard

scorecard <- tibble(
  model        = c("prior_year_avg", "same_dow_mean", "50/50 blend"),
  train_data   = "n/a — no fitting",
  val_wmape    = c(wmape_val_b1,  wmape_val_b2,  wmape_val_b3),
  test_wmape   = c(wmape_test_b1, wmape_test_b2, wmape_test_b3)
) |>
  mutate(
    val_wmape_pct  = percent(val_wmape,  accuracy = 0.01),
    test_wmape_pct = percent(test_wmape, accuracy = 0.01)
  )

scorecard |>
  select(model, train_data, val_wmape_pct, test_wmape_pct) |>
  print()
```

```{r}
#| label: scorecard-plot
#| fig-width: 9
#| fig-height: 4

scorecard |>
  pivot_longer(c(val_wmape, test_wmape),
               names_to = "split", values_to = "wmape") |>
  mutate(
    split = if_else(split == "val_wmape", "Val (Sep 2025)", "Test (Oct-Dec 2025)"),
    model = fct_reorder(model, wmape)
  ) |>
  ggplot(aes(x = model, y = wmape, fill = split)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = percent(wmape, accuracy = 0.01)),
            position = position_dodge(width = 0.6),
            vjust = -0.4, size = 3.2) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Val (Sep 2025)" = "#3B82C4",
                                "Test (Oct-Dec 2025)" = "#6B7FA3")) +
  labs(
    title = "Baseline wMAPE — Val & Test",
    subtitle = "These are the numbers every downstream model must beat",
    x = NULL, y = "wMAPE", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        legend.position    = "top")
```

---

## 7. Error Breakdown — Val Set

Understand where the baseline fails before reaching for a model.

```{r}
#| label: error-by-category
#| fig-width: 10
#| fig-height: 4

val_b1 |>
  filter(!is.na(quantity), quantity > 0) |>
  group_by(category) |>
  summarise(
    wmape  = wmape(quantity, predicted),
    volume = sum(quantity),
    .groups = "drop"
  ) |>
  mutate(
    category = fct_reorder(category, wmape),
    flag     = wmape > wmape_val_b1 + 0.01
  ) |>
  ggplot(aes(x = category, y = wmape, fill = flag)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = wmape_val_b1, linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_fill_manual(values = c("FALSE" = "#3B82C4", "TRUE" = "#E05C5C"),
                    guide = "none") +
  labs(
    title    = "Baseline 1 (prior_year_avg) wMAPE by category — Val",
    subtitle  = "Red = more than 1pp above overall baseline",
    x = NULL, y = "wMAPE"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())
```

```{r}
#| label: error-by-restaurant
#| fig-width: 10
#| fig-height: 4

val_b1 |>
  filter(!is.na(quantity), quantity > 0) |>
  group_by(restaurant_id) |>
  summarise(
    wmape  = wmape(quantity, predicted),
    volume = sum(quantity),
    .groups = "drop"
  ) |>
  mutate(
    restaurant_id = fct_reorder(restaurant_id, wmape),
    flag          = wmape > wmape_val_b1 + 0.01
  ) |>
  ggplot(aes(x = restaurant_id, y = wmape, fill = flag)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = wmape_val_b1, linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_fill_manual(values = c("FALSE" = "#3B82C4", "TRUE" = "#E05C5C"),
                    guide = "none") +
  labs(
    title    = "Baseline 1 (prior_year_avg) wMAPE by restaurant — Val",
    subtitle  = "Red = more than 1pp above overall baseline",
    x = NULL, y = "wMAPE"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())
```

---

## 8. Best Baseline Submission

Write the best-performing baseline as a valid submission file in case
it is needed as a fallback.

```{r}
#| label: write-submission

# Use whichever baseline scored best on val
best_baseline <- list(b1 = wmape_val_b1, b2 = wmape_val_b2, b3 = wmape_val_b3)
best_name     <- names(which.min(unlist(best_baseline)))

best_preds <- switch(best_name,
  b1 = test_b1,
  b2 = test_b2,
  b3 = test_b3
)

submission_baseline <- best_preds |>
  select(date, restaurant_id, menu_item_id, predicted_quantity = predicted)

stopifnot(nrow(submission_baseline) == 69000)

write_csv(submission_baseline,
          glue("data/submission_baseline_{best_name}.csv"))

glue("Saved: data/submission_baseline_{best_name}.csv ({nrow(submission_baseline)} rows)")
glue("Best baseline val wMAPE: {percent(min(unlist(best_baseline)), accuracy = 0.01)}")
```

---

## Summary

```{r}
#| label: final-summary

scorecard |>
  select(model, val_wmape_pct, test_wmape_pct) |>
  arrange(val_wmape_pct) |>
  print()

cat("\nTarget to beat in 02_lightgbm.qmd: val wMAPE <",
    percent(min(c(wmape_val_b1, wmape_val_b2, wmape_val_b3)), accuracy = 0.01), "\n")
```
