---
title: "02 — LightGBM"
author: "Mario"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(tidymodels)
library(bonsai)
library(lightgbm)
library(scales)
library(glue)
library(patchwork)
```

---

## Purpose

Fits LightGBM against **all three training variants** from
`00_data_prep.qmd` in a single render. No manual commenting or
uncommenting required — a loop handles everything.

All three variants share identical features, model spec, wMAPE function,
and val/test splits. Any difference in wMAPE is attributable to the null
strategy alone.

---

## 1. Load Config, Val & Test

Val and test never change across variants — load them once here.

```{r}
#| label: load-fixed-splits

cfg  <- readRDS("data/prep_config.rds")
val  <- read_csv("data/val.csv",  show_col_types = FALSE)
test <- read_csv("data/test.csv", show_col_types = FALSE)

FEATURES <- cfg$FEATURES
TARGET   <- cfg$TARGET

CAT_COLS <- c("restaurant_id", "menu_item_id", "category",
              "restaurant_name", "city", "state")

glue("Val rows : {nrow(val)}")
glue("Test rows: {nrow(test)}")
glue("Features : {length(FEATURES)}")
```

---

## 2. wMAPE Helper

```{r}
#| label: wmape-helper

wmape <- function(actual, predicted) {
  mask <- !is.na(actual) & actual > 0
  sum(abs(actual[mask] - predicted[mask])) / sum(actual[mask])
}

score_wmape <- function(df) {
  wmape(df[[TARGET]], df$predicted)
}
```

---

## 3. Baseline Reference

Computed once — identical regardless of training variant.

```{r}
#| label: baseline-reference

val_baseline_wmape <- val |>
  mutate(predicted = coalesce(prior_year_avg, combo_mean) |> pmax(0)) |>
  score_wmape()

test_baseline_wmape <- test |>
  mutate(predicted = coalesce(prior_year_avg, combo_mean) |> pmax(0)) |>
  score_wmape()

glue("Baseline (prior_year_avg) — Val: {percent(val_baseline_wmape, accuracy = 0.01)} | Test: {percent(test_baseline_wmape, accuracy = 0.01)}")
```

---

## 4. Model Spec

Defined once, reused for every variant.

```{r}
#| label: model-spec

lgbm_spec <- boost_tree(
  trees          = 1000,
  tree_depth     = 6,
  min_n          = 80,
  learn_rate     = 0.04,
  loss_reduction = 0.2,
  sample_size    = 0.9,
  mtry           = 0.85
) |>
  set_engine(
    "lightgbm",
    objective    = "regression_l1",
    num_leaves   = 127,
    lambda_l2    = 0.8,
    bagging_freq = 1,
    seed         = 42,
    num_threads  = 0,
    verbosity    = -1,
    counts       = FALSE
  ) |>
  set_mode("regression")
```

---

## 5. Fit All Three Variants

Each iteration loads training data → builds recipe → log-transforms
target → fits model → predicts val and test → scores wMAPE → saves
submission CSV. Results accumulate into `all_results`.

```{r}
#| label: fit-all-variants

VARIANTS <- tribble(
  ~label,          ~path,
  "dropped",       "data/train_dropped.csv",
  "imputed_combo", "data/train_imputed_combo_median.csv",
  "imputed_dow",   "data/train_imputed_dow_median.csv"
)

all_results    <- tibble()
all_val_scored <- tibble()
best_model_fit <- NULL   # holds dropped-nulls fit for diagnostics

for (i in seq_len(nrow(VARIANTS))) {

  v_label <- VARIANTS$label[i]
  v_path  <- VARIANTS$path[i]

  cat(glue("\n── Variant {i}/3: {v_label} ──────────────────────────\n"))

  # ── Load ─────────────────────────────────────────────────────────────────
  train <- read_csv(v_path, show_col_types = FALSE)
  cat(glue("   Train rows : {nrow(train)}\n"))

  # ── Recipe ───────────────────────────────────────────────────────────────
  lgbm_recipe <- recipe(
      as.formula(paste(TARGET, "~ .")),
      data = train |> select(all_of(c(TARGET, FEATURES, CAT_COLS)))
    ) |>
    step_mutate(across(all_of(CAT_COLS), as.factor))
    # Note: step_impute_median intentionally removed — LightGBM handles NaN
    # natively via use_missing=TRUE (default). NaN in lag_364 means
    # "less than one year of history" — that is signal, not noise.

  lgbm_prep   <- prep(lgbm_recipe, training = train)
  train_baked <- bake(lgbm_prep, new_data = train)
  val_baked   <- bake(lgbm_prep, new_data = val)
  test_baked  <- bake(lgbm_prep, new_data = test)

  # ── Log-transform target + volume-aware sample weights ──────────────────
  # Upweight high-volume rows so the model optimises closer to wMAPE.
  # Weight formula mirrors the Python benchmark: 1 + sqrt(quantity).
  # Using fit_xy() to keep case_wt separate from the predictor matrix —
  # the formula interface (quantity ~ .) would treat case_wt as a feature.
  train_log <- train_baked |>
    mutate(quantity = log1p(quantity))

  x_train <- train_log |> select(-all_of(TARGET))
  y_train <- train_log[[TARGET]]
  wts     <- hardhat::importance_weights(1 + sqrt(pmax(y_train, 0)))

  # ── Fit ──────────────────────────────────────────────────────────────────
  cat(glue("   Fitting...\n"))
  lgbm_fit <- lgbm_spec |>
    fit_xy(x = x_train, y = y_train, case_weights = wts)
  cat(glue("   Done.\n"))

  # ── Predict & back-transform ─────────────────────────────────────────────
  val_preds <- predict(lgbm_fit,
                       new_data = val_baked |> select(-all_of(TARGET))) |>
    mutate(.pred = pmax(expm1(.pred), 0))

  test_preds <- predict(lgbm_fit,
                        new_data = test_baked |> select(-all_of(TARGET))) |>
    mutate(.pred = pmax(expm1(.pred), 0))

  val_scored  <- val  |> mutate(predicted = val_preds$.pred,  variant = v_label)
  test_scored <- test |> mutate(predicted = test_preds$.pred, variant = v_label)

  # ── Score ─────────────────────────────────────────────────────────────────
  wm_val  <- score_wmape(val_scored)
  wm_test <- score_wmape(test_scored)

  cat(glue("   Val wMAPE : {percent(wm_val,  accuracy = 0.01)}\n"))
  cat(glue("   Test wMAPE: {percent(wm_test, accuracy = 0.01)}\n"))

  # ── Change 5: Specials override — replace with seasonal baseline if it helps
  # Specials have intermittent, event-driven demand where lags are weak anchors.
  # Test on val first — only apply to test if val wMAPE actually improves.
  val_override <- val_scored |>
    mutate(predicted = if_else(
      category == "Specials",
      coalesce(prior_year_avg, combo_mean) |> pmax(0),
      predicted
    ))
  wm_val_override <- score_wmape(val_override)
  cat(glue("   Val wMAPE with Specials override: {percent(wm_val_override, accuracy = 0.01)}\n"))

  if (wm_val_override < wm_val) {
    test_scored <- test_scored |>
      mutate(predicted = if_else(
        category == "Specials",
        coalesce(prior_year_avg, combo_mean) |> pmax(0),
        predicted
      ))
    wm_test <- score_wmape(test_scored)
    wm_val  <- wm_val_override
    cat(glue("   ✓ Specials override applied — new val wMAPE: {percent(wm_val, accuracy = 0.01)}\n"))
  } else {
    cat(glue("   Specials override did not help val — skipped.\n"))
  }

  # ── Collect ───────────────────────────────────────────────────────────────
  all_results <- bind_rows(
    all_results,
    tibble(
      variant         = v_label,
      train_rows      = nrow(train),
      val_wmape       = wm_val,
      test_wmape      = wm_test,
      val_vs_baseline = wm_val - val_baseline_wmape
    )
  )
  all_val_scored <- bind_rows(all_val_scored, val_scored)

  # ── Save submission ───────────────────────────────────────────────────────
  submission <- test_scored |>
    select(date, restaurant_id, menu_item_id,
           predicted_quantity = predicted) |>
    mutate(date               = as.character(date),
           predicted_quantity = round(predicted_quantity, 4)) |>
    arrange(date, restaurant_id, menu_item_id)

  stopifnot("Submission must be 69,000 rows" = nrow(submission) == 69000)
  out_path <- glue("data/submission_lgbm_{v_label}.csv")
  write_csv(submission, out_path)
  cat(glue("   Saved: {out_path}\n"))

  # keep dropped fit for diagnostics
  if (v_label == "dropped") best_model_fit <- lgbm_fit
}

cat("\n── All variants complete ─────────────────────────────────\n")
```

---

## 6. Comparison Scorecard

```{r}
#| label: scorecard

scorecard <- bind_rows(
  tibble(
    variant         = "baseline (prior_year_avg)",
    train_rows      = NA_integer_,
    val_wmape       = val_baseline_wmape,
    test_wmape      = test_baseline_wmape,
    val_vs_baseline = 0
  ),
  all_results
) |>
  mutate(
    val_pct        = percent(val_wmape,        accuracy = 0.01),
    test_pct       = percent(test_wmape,        accuracy = 0.01),
    vs_baseline_pp = percent(val_vs_baseline,   accuracy = 0.01)
  )

scorecard |>
  select(variant, train_rows, val_pct, test_pct, vs_baseline_pp) |>
  print()
```

```{r}
#| label: scorecard-plot
#| fig-width: 10
#| fig-height: 4

scorecard |>
  filter(variant != "baseline (prior_year_avg)") |>
  pivot_longer(c(val_wmape, test_wmape),
               names_to = "split", values_to = "wmape") |>
  mutate(
    split   = if_else(split == "val_wmape",
                      "Val (Sep 2025)", "Test (Oct-Dec 2025)"),
    variant = fct_reorder(variant, wmape)
  ) |>
  ggplot(aes(x = variant, y = wmape, fill = split)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_hline(yintercept = val_baseline_wmape,
             linetype = "dashed", color = "grey40", linewidth = 0.8) +
  geom_text(aes(label = percent(wmape, accuracy = 0.01)),
            position = position_dodge(width = 0.6),
            vjust = -0.4, size = 3.2) +
  annotate("text", x = 0.55, y = val_baseline_wmape + 0.003,
           label = glue("Baseline {percent(val_baseline_wmape, accuracy = 0.01)}"),
           hjust = 0, size = 3, color = "grey40") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Val (Sep 2025)"      = "#3B82C4",
                                "Test (Oct-Dec 2025)" = "#6B7FA3")) +
  labs(
    title    = "LightGBM wMAPE — all three training variants",
    subtitle = "Dashed line = prior_year_avg baseline",
    x = NULL, y = "wMAPE", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        legend.position    = "top")
```

---

## 7. Feature Importance

Uses the `dropped` variant — rankings are identical across all three.

```{r}
#| label: feature-importance
#| fig-width: 9
#| fig-height: 8

best_model_fit$fit |>
  lightgbm::lgb.importance(percentage = TRUE) |>
  as_tibble() |>
  slice_max(Gain, n = 25) |>
  mutate(Feature = fct_reorder(Feature, Gain)) |>
  ggplot(aes(x = Gain, y = Feature)) +
  geom_col(fill = "#3B82C4", width = 0.7) +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title    = "LightGBM feature importance — dropped variant",
    subtitle = "Top 25 features by gain",
    x = "Gain (%)", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

---

## 8. Diagnostics — Val Set (dropped variant)

```{r}
#| label: residual-plots
#| fig-width: 12
#| fig-height: 5

val_diag <- all_val_scored |>
  filter(variant == "dropped", !is.na(quantity), quantity > 0) |>
  mutate(
    abs_pct_error = abs(quantity - predicted) / quantity,
    residual      = quantity - predicted
  )

p1 <- val_diag |>
  ggplot(aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_hline(yintercept = 0, color = "red", linewidth = 0.8) +
  labs(title = "Residuals vs Predicted — Val (dropped)",
       x = "Predicted", y = "Residual") +
  theme_minimal(base_size = 11)

p2 <- val_diag |>
  mutate(pct_error_capped = pmin(abs_pct_error, 1)) |>
  ggplot(aes(x = pct_error_capped)) +
  geom_histogram(bins = 50, fill = "#3B82C4", color = "white") +
  scale_x_continuous(labels = percent_format()) +
  labs(title = "Distribution of |% Error| — Val (capped at 100%)",
       x = "Absolute % Error", y = "Count") +
  theme_minimal(base_size = 11)

p1 + p2

cat("Median |% error|  :", percent(median(val_diag$abs_pct_error), accuracy = 0.1), "\n")
cat("90th pct |% error|:", percent(quantile(val_diag$abs_pct_error, 0.9), accuracy = 0.1), "\n")
```

```{r}
#| label: wmape-by-category
#| fig-width: 10
#| fig-height: 4

best_val_wmape <- all_results |>
  filter(variant == "dropped") |>
  pull(val_wmape)

all_val_scored |>
  filter(variant == "dropped", !is.na(quantity), quantity > 0) |>
  group_by(category) |>
  summarise(wmape  = wmape(quantity, predicted),
            volume = sum(quantity), .groups = "drop") |>
  mutate(category = fct_reorder(category, wmape),
         flag     = wmape > best_val_wmape + 0.01) |>
  ggplot(aes(x = category, y = wmape, fill = flag)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = best_val_wmape,
             linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_fill_manual(values = c("FALSE" = "#3B82C4", "TRUE" = "#E05C5C"),
                    guide = "none") +
  labs(title = "wMAPE by category — LightGBM (dropped) Val",
       subtitle = "Red = more than 1pp above overall",
       x = NULL, y = "wMAPE") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())
```

```{r}
#| label: wmape-by-restaurant
#| fig-width: 10
#| fig-height: 4

all_val_scored |>
  filter(variant == "dropped", !is.na(quantity), quantity > 0) |>
  group_by(restaurant_id) |>
  summarise(wmape  = wmape(quantity, predicted),
            volume = sum(quantity), .groups = "drop") |>
  mutate(restaurant_id = fct_reorder(restaurant_id, wmape),
         flag          = wmape > best_val_wmape + 0.01) |>
  ggplot(aes(x = restaurant_id, y = wmape, fill = flag)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = best_val_wmape,
             linetype = "dashed", color = "grey40") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_fill_manual(values = c("FALSE" = "#3B82C4", "TRUE" = "#E05C5C"),
                    guide = "none") +
  labs(title = "wMAPE by restaurant — LightGBM (dropped) Val",
       subtitle = "Red = more than 1pp above overall",
       x = NULL, y = "wMAPE") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())
```

---

## 9. Best Submission

Automatically selects the variant with the lowest val wMAPE.

```{r}
#| label: best-submission

best_variant <- all_results |>
  slice_min(val_wmape, n = 1) |>
  pull(variant)

best_val  <- all_results |> filter(variant == best_variant) |> pull(val_wmape)
best_test <- all_results |> filter(variant == best_variant) |> pull(test_wmape)
best_path <- glue("data/submission_lgbm_{best_variant}.csv")

cat(glue("Best variant  : {best_variant}\n"))
cat(glue("Val wMAPE     : {percent(best_val,  accuracy = 0.01)}\n"))
cat(glue("Test wMAPE    : {percent(best_test, accuracy = 0.01)}\n"))
cat(glue("Submission    : {best_path}\n"))

# Confirm 69,000 rows
final_sub <- read_csv(best_path, show_col_types = FALSE)
stopifnot("Must be 69,000 rows" = nrow(final_sub) == 69000)
cat(glue("Row count     : {nrow(final_sub)} ✓\n"))
```
